## NCAA Football Attendance

Let's load the tidyverse.

```{r}
library(tidyverse)
```

And read in the attendance data:

```{r}
attendance <- read_csv("https://dwillis.github.io/sports-data-files/attendance.csv")
```

Using the tidyverse's `pivot_longer` function, turn the currently wide data into longer data and save it as a new dataframe called `attendance_longer`:

```{r}
attendance_longer <- attendance |> pivot_longer(cols = starts_with("20"), names_to = "Year", values_to = "Attendance")

attendance_longer |> filter(Attendance >= 500000) |>
  group_by((Institution)) |>
summarize (years = n()) |>
  filter(years == 1)
```

Let's try to figure out whether football attendance has returned to "normal" by 2023, defining "normal" as 2019 figures. For that we'll need to isolate those two years using `filter()` and get rid of any NAs:

```{r}
attendance_two_years <- attendance_longer |> filter(Year %in% c(2019, 2023)) |> na.omit()
```

Now, in order to calculate a percentage change between those two years, we'll need to go back to wider data and again ditch teams where we don't have both years:

```{r}
attendance_wide <- attendance_two_years |> pivot_wider(names_from = Year, values_from = Attendance) |> na.omit()
```

Finally, we'll add a percentage change column to that dataframe (remember, it's (New-Old)/Old) and arrange it to see the results.

```{r}
attendance_wide <- attendance_wide |> mutate(percent = (`2023`-`2019`)/`2019`*100)
```

Some of these clearly are influenced by performance on the field. Below, describe the most interesting results (positive or negative) and what you think might be going on with them. Feel free to do some googling and describe what other data would be helpful to clarify what the best stories here might be. What surprises you about these results?

**ANSWER**
A pretty intresting finding was UNLV's increase in attendance. UNLV's full name is the University of Nevada Las Vegas, and has a stadium on campus that is renovated but chose to move to Allegiant Stadium in 2020. This is the biggest change from 2019 to 2023 with the only difference being UNLV being good in Mountain West play in 2023. So this move could have the numbers to show that it was a good choice.
```{r}
B <- read.csv("https://dwillis.github.io/sports-data-files/big_ten_no_2020.csv")
```


```{r}
big_ten_avg <- B |>
  group_by(Institution) |>
  summarize(
    avg_attendance = mean(Attendance) ,
    avg_wins = mean(Wins),
    sd_attendance = sd(Attendance),
    sd_wins = sd(Wins)
  ) |>
  arrange(desc(sd_wins))

big_ten_avg

```
Nebraska's fanbase continues to pack crowds despite averaging under six wins a season. They rank fourth in the Big Ten in attendance despite Wisconsin finishing with over 4 wins a season more than Nebraska. Same situation with Iowa. It's not like these other programs dont have impressive attendance numbers either. 
```{r}
library(tidyverse)
attendance <- read_csv('https://dwillis.github.io/sports-data-files/attendance.csv')
```
```{r}
attendance |> 
  arrange(desc(`2023`)) |> 
  top_n(10) |> 
  select(Institution, `2023`)
```
```{r}
top10 <- attendance |>
  arrange(desc(`2023`)) |> 
  top_n(10) |> 
  select(Institution, `2023`)
```
```{r}
ggplot() + 
  geom_bar(
    data=top10, 
    aes(x=Institution)
  )
```

```{r}
ggplot() + 
  geom_bar(
    data=top10, 
    aes(x=Institution, weight=`2023`)
  )
```

```{r}
ggplot() + 
  geom_bar(
    data=top10, 
    aes(
      x=reorder(Institution, `2023`), 
      weight=`2023`
      )
    )
```
```{r}
ggplot() + 
  geom_bar(
    data=top10, 
    aes(
      x=reorder(Institution, -`2023`), 
      weight=`2023`
      )
    )
```

```{r}
library(scales)
ggplot() + 
  geom_bar(
    data=top10, 
    aes(
      x=reorder(Institution, -`2023`), 
      weight=`2023`
      )
    ) + 
  scale_y_continuous(labels=comma)
```
```{r}
ggplot() + 
  geom_bar(
    data=top10, 
    aes(
      x=reorder(Institution, -`2023`), 
      weight=`2023`)
    ) + 
  scale_y_continuous(labels=comma) + 
  labs(
    title="Top 10 Football Programs By Attendance", 
    x="School", 
    y="Attendance"
)
```
```{r}
ggplot() + 
  geom_bar(
    data=top10, 
    aes(x=reorder(Institution, -`2023`),
        weight=`2023`)) + 
  scale_y_continuous(labels=comma) + 
  labs(
    title="Top 10 Football Programs By Attendance", 
    x="School", 
    y="Attendance") + 
  theme_light()
```
```{r}
ggplot() + 
  geom_bar(
    data=top10, 
    aes(x=reorder(Institution, -`2023`),
        weight=`2023`)) + 
  scale_y_continuous(labels=comma) + 
  labs(
    title="Top 10 Football Programs By Attendance", 
    x="School", 
    y="Attendance") + 
  theme_minimal() + 
  coord_flip()
```

