The goal of this presentation is to find the best players on the worst teams, who has succeeded the most but cannot reap the benefits. For this, I want all players in the past 20 years that is a qualified hitter, which Fangraphs filters out themselves. This is what I have from that. 
```{r}
install.packages("baseballr")
if (!requireNamespace('pacman', quietly = TRUE)){
  install.packages('pacman')
}
pacman::p_load_current_gh("BillPetti/baseballr")

```
```{r}

install.packages("mlbstatsR")
```

```{r}
library(baseballr)
library(dplyr)
library(tidyverse)
library(readr)
library(formattable)


```
```{r}
batting_data <- read_csv("C:/desktop/jour479x_fall_2024/JoniacoBaseball - Sheet2.csv")

```
This is all qualified batters from 2004 to 2024, there are a couple problems with this set. Some players have played on two teams. I want only players that stuck out the whole season. So I'm filtering for only players that have these teams. 
```{r}
Teams <- c("ARI", "LAD", "NYY", "BAL", "MIL", "PHI", "SDP", "HOU", "NYM", "MIN", "TOR", "CHC", "CLE", "SEA", "BOS", "KCR", "DET", "ATL", "SFG", "TEX", "STL", "TBR", "OAK", "WSN","WSH", "CIN", "PIT", "MIA", "LAA", "COL", "CHW","FLA","TBD")
filtered <- batting_data |> filter(Team %in% Teams) |> filter (Season %in% c(2004,2005,2006,2007,2008,2009,2010,2011,2012,2013,2014,2015,2016,2017,2018,2019,2021,2022,2023,2024))
```
The main statstic I will be using is Wins Above Replacement, or WAR. This is an estimate of how many wins a player is worth over a replacement level player. It has a very convluteted measurement, but for the sake of this presentation, I'm taking it at face value for wins. 

Here is a table of Games Played and fangraphs wins above replacement (FWAR) accumulated, and arranged by WAR divided by Games played. This is a preliminary idea of what we're looking at. Barry Bonds' 2004 season is well regarded as one of the best of all time. By looking at this stat, in 147 games, he was on pace for over 12 Wins above replacement. WAR/G also appreciates players who had injuries at some point, such as 2018 Mookie Betts or 2018 Mike Trout, both who missed over 20 games and not at the Top in the fWAR department, but high in the ratio to WAR to Games played.
```{r}
filtered |> select(Season,Name,Team,G,WAR) |> mutate(WAR/G) |> arrange(desc(WAR/G))
```
The idea is to create a metric, divding it by the WAR of each player with the wins of the team they were from, there are other ways to find production, like looking at wRC+ for offensive production or Defensive Runs Saved or defensive grade for defense, but for now, WAR/Wins is the start.

WAR/Wins, or WARSHARE is a metric that takes the total acculmated WAR in a season and divides it by the amount of wins a team had in a given season. So if a player has 10 WAR, and the team wins 100 Games, the WARSHARE is 0.10, or 10%. fWAR is not a perfect statistic, but it is a good baseline to see who is most valuable. 

First you need to load in the Wins from each team from 2004 to 2024, this is downloaded on my computer as of now and I will find a way to make this readily available.
Then you Bind together the Season from the standings, and the Team. Both tables have these values so binding it makes it easy.

```{r}
standings <- read_csv("C:/desktop/jour479x_fall_2024/presentations/Standi.csv")
merged_data <- merge(filtered, standings, by = c("Season", "Team"), all.x = TRUE)

# Rename the Wins column if needed
colnames(merged_data)[colnames(merged_data) == "W"] <- "Team_Wins"
```
With this data we want the Season, Name, Team, Games Played, Team Wins and add WAR/Team Wins. I also want to add an inqury of players with over 10% of their teams wins according to WAR and put a parameter on them to highlight this later. 

```{r}
combined <- merged_data |> select(Season,Name,Team,G,WAR,Team_Wins) |> mutate(WARSHARE = WAR/Team_Wins, Highlight = ifelse(WARSHARE > 0.10, "High", "Low")) |> filter((!is.na(Team_Wins))) |> arrange(desc(WARSHARE)) 
combined
```
Here is a scatterplot of the data, showing Team Wins over WAR. There are a couple of standouts. 
```{r}
ggplot(data = combined, aes(x = Team_Wins, y = WAR)) +
  geom_point(size = 1) + 
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
   geom_point(aes(color = Highlight)) +
  labs(x = "Team Wins", y = "WAR") +
  theme_minimal() +
  theme(legend.title = element_blank()) 


```
Notice the trendline in the scatterplot. In this over 2,800 player sample, if your team is under 60 wins you most likely will be around a 2 WAR player. If your team has over 100 Wins, you most likely will be a 4 WAR player. 
```{r}
Over10 <- combined |> filter(Highlight == "High")
Over10 |> group_by(Name) |> summarize(Count = n()) |> arrange(desc(Count))
```
Since 2004, a player has made up over 10% WAR ShareR 27 times. Mike Trout himself has done it six times. 
```{r}
ggplot(data = combined, aes(x = Team_Wins, y = WAR)) +
  geom_point(size = 1) +  # Points for all players, adjust size as needed
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
   geom_point(aes(color = Highlight)) +
  geom_point(data = subset(combined, Name == "Mike Trout"), aes(color = "Mike Trout"), size = 2) +  # Highlight Mike Trout
  labs(x = "Team Wins", y = "WAR") +
  theme_minimal() +
  theme(legend.title = element_blank())  # Remove legend title if desired

```
The gap between Mike Trout and the rest of the MLB in this sample is truly remarkable, well until you remember 2004 Barry Bonds. But other than those two seasons, nothing is even close to them production wise.
```{r}
# Calculate z-scores
zscores <- combined %>%
  mutate(zscore = as.numeric(scale(WARSHARE, center = TRUE, scale = TRUE)))
zscores
```
This sample is all qualified batters, so players who would be eligible for awards based on play time. Some players seasons are missing from this sample due to the fact they missed too much time. This would also draw the WAR average down, WAR is a culmuative stat, so if you are an above replacement level player, it goes up the more you play games. 


```{r}


combined |> group_by(Name) |> summarize(Total = mean(WARSHARE)*100,Seasons = n()) |> arrange(desc(Seasons))

```
Adding one more parameter to this statistic. Because qualified batters is based on plate appereances, they are able to miss some games and still be considered qualified. Well do players who produce at a high rate but missed some games maybe get boosted if we try to account for missed time? 
```{r}
combined |> mutate(Adjust = `WARSHARE`/(G/162)) |> arrange(desc(Adjust)) 
```
If we adjust for games missed, in games he played, Mike Trout accounted for over 13% of his team's wins four times. Other than 2004 Barry Bonds, no other player has done that once. 
```{r}
combined |> mutate(Adjust = `WARSHARE`/(G/162)) |> arrange(desc(Adjust)) |> mutate(Difference = Adjust - `WARSHARE`) |> arrange(desc(Difference))
```
Finally, look at some of the players who were on huge paces, but missed out on too many games to crack the top of the leader board. Joe Mauer had 5 WAR in 113 games on a team that would go on to win 66 games that year. If he stayed healthy for 162 games, he'd be in a select club of players that were worth over 10% of their wins according to WAR in the past 20 years. Chipper Jones in 2008 missed over 40 games, yet stil was close to 10% of his teams WAR. If he played 162 games, he'd have one of the best seasons in the past 20 years. 

So we know about Mike Trout, but what about players in that left hand quadrant of the scatterplot? Subpar team wins but good performances.Setting the parameters at 5 WAR, a very solid player, and less than 60 wins, which is really bad. 
```{r}
combined |> filter(Team_Wins <= 60, WAR >= 5)
```
Only Six guys show up, all players who would go on to be All-Stars at some point in their career on terrible teams. Cedric Mullins' 6 WAR is impressive, but his 11% WAR share makes it almost as impressive as some of the MVP season's seen above.

This is the extreme, all of these players were having All-Star season's on bad teams. What about players that started on bad teams and were able to improve? From here I'm going to define the quartiles of good and bad teams. Here's the 
```{r}
summary(combined)
```
I want to seprate players into quartiles from team wins, But for now we'll focus on the last quadrant of players on teams with less than 74 wins. 
```{r}
FQ <- combined |> filter(Team_Wins <= 74)
Med <- combined |> filter(Team_Wins >= 74 & Team_Wins <= 84)
TQ <- combined |> filter(Team_Wins >= 84 & Team_Wins <= 92)
Mx <- combined |> filter(Team_Wins >= 92)

```
```{r}
summary(FQ)
```
A quick summary look, the Mean WAR is 2.2, more than half a game less than the total dataset. 
What does the average WAR look like for good teams vs. bad teams? Working off the quadrants shown from the combined dataset, the first quadrant are players on teams with less than 74 wins, where as the last quadrant are players on teams above 92 wins. How do these compare? Here's a boxplot to show the averages and the outliers. 

Simliar to the line chart, the median WAR of players who are on teams with less than 74 wins is less than the first quartile of players on teams with over 92 wins. This seems obvious, but it puts into perspective our outlier group at the top of the graph. 
```{r}
FQ |> arrange(desc(WARSHARE))
```
Looking at this table we see familiar names like Mike Trout, Cedric Mullins. But what about players that aren't All-Star caliber but are good? The third quartile of players under 74 wins is around 3.4 WAR, round up to 4WAR and lets see who remains.

```{r}
SweetSpot <- FQ |> filter(WAR <= 4 & WAR >= 2.2) |> arrange(desc(WAR))
SweetSpot
```
At this point in the sample, most players have simliar WARs and WARSHARES. The parameters of this data is between a 2 and 4 WAR and a team under 74 wins. But there's still around 217 players who fit this criteria. At the top is David DeJesus, who stands out of this sample by missing 40 games and being over 7% of his teams wins in WARSHARE. 
```{r}
combined |> filter(Name == "David DeJesus") |> arrange(Season)
```
Not shown is DeJesus' rookie year, where he finished 6th in Rookie of the Year voting in 2004. His 2005 season was a successful sophmore season, where it looked like he was a future All-Star for the Royals. Looking at his career Trajectory, DeJesus had 7 qualified seasons, all on teams with 75 wins or less. His tragetory peaked in his second season, where he had 4.2 WAR in 119 games, a pretty impressive feat. It took DeJesus 8 years before he got to a team with over 80 wins, but by then, his production wasn't there. 
```{r}
library(ggplot2)
dejesus_data <- combined |> 
  filter(Name == "David DeJesus") 
ggplot(dejesus_data, aes(x = Season, y = WAR)) +
  geom_line(color = "blue") +
  geom_point(color = "red") +  # Optional: add points to highlight individual seasons
  labs(title = "David DeJesus Qualified Seasons", 
       x = "Season", y = "WAR") +
  theme_minimal()

```
The goal of this project is to find players who were able to succeed on bad teams, while DeJesus did in his first two years, the rest of his career he didn't see simliar success, well what players did? 

Here is a graph of season's since the sweet spot season. If a player was on a team below 74 Wins and in the sweet spot, this is a plot of the trajectory. 


```{r}
first_season_benchmark <- ss |>
  group_by(Name) |>
  summarize(First_Season = min(Season), .groups = "drop")
progression_data <- combined |>
  filter(Name %in% first_season_benchmark$Name) |> 
  left_join(first_season_benchmark, by = "Name") |>
  arrange(Name, Season) |> 
  mutate(Seasons_Since = Season - First_Season) |> 
  select(-First_Season) |> filter(Seasons_Since >= 0) 
pd <- progression_data |>
  group_by(Name) |>
  arrange(Name, Season) |>
  mutate(keep = ifelse(Seasons_Since == 0 & lag(Seasons_Since == 0, default = FALSE), FALSE, TRUE)) |>  # Mark subsequent 0s as FALSE
  filter(keep) |>  # Keep only rows marked as TRUE
  select(-keep) |>  # Remove helper column
  ungroup()

comp <- pd |>
  group_by(Name) |>
  arrange(Season) |>
  mutate(totalWAR = cumsum(WAR)) |>
  ungroup() |> arrange(Name)
comp
jones <- comp|> filter(Name == "Adam Jones")
jesus <- comp |> filter(Name == "David DeJesus")
```

```{r}
ggplot(comp, aes(x = Seasons_Since, y = WAR, group = Name)) +
  geom_line() +
  labs(title = "Player WAR Progression Over Time",
       x = "Seasons Since First Appearance",
       y = "WAR") +
  theme_minimal()
```

```{r}
comp |>
  filter(Team_Wins < 74) |>  # Filter for teams with fewer than 74 wins
  group_by(Name) |>  # Group by player name
  summarize(Count= n()) |>  # Count occurrences for each player
  ungroup()  |> arrange(desc(Count))
```
```{r}
comp |> arrange(desc(WAR))
```
```{r}
library(ggplot2)
comp <- comp |>
  group_by(Name) |>
  mutate(High_WAR_Player = any(WAR > 7)) |>
  ungroup()

# Plot with highlighted lines and points
ggplot(comp, aes(x = Seasons_Since, y = WAR, group = Name)) +
  geom_line(aes(color = High_WAR_Player), size = 1) +  # Highlight lines for high-WAR players
  geom_point(data = comp |> filter(WAR > 7), aes(x = Seasons_Since, y = WAR), color = "black", size = 1.5) +  # Highlight points for WAR > 7
   geom_hline(yintercept = 7, linetype = "dashed", color = "darkred", size = 0.8) + 
  scale_color_manual(values = c("FALSE" = "gray", "TRUE" = "red")) +  # Blue for high-WAR players, gray otherwise
  labs(title = "Player WAR Progression Over Time",
       x = "Seasons Since First Appearance",
       y = "WAR") +
  theme_minimal()
```

